Competing Constructions for German Measure Noun Phrases

### 2017-12-21 Latex/article.tex
More reviewers' comments.

\documentclass[USenglish]{article}

@@

\usepackage[small]{dgruyter}
\usepackage{microtype}
\usepackage{enumerate}
\usepackage{eqnarray}
\usepackage{colortbl}

@@

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer. Submitted April 2017. Do not quote.}
  %\affil[1]{}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{In this paper, a case alternation in German measure noun phrases is examined from the perspective of cognitively oriented corpus linguistics.
    In certain pseudo-partitive constructions, the embedded kind-denoting noun either agrees in its case with the measure head noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    My analysis is formulated in terms of Prototype Theory.
    I assume that the choice of alternants is influenced by non-alternating neighbouring constructions representing the prototypes more directly.
    I argue that the frequencies with which individual lemmas occur in these prototypical non-alternating constructions partially predict which alternant is chosen and that the genitive alternant is prototypical of strongly grammaticalised measure nouns.
    I present a large-scale corpus study using the DECOW corpus to support this theory.
    In the statistical analysis, I compare Maximum Likelihood estimators to Bayesian estimators (recently proposed for similar studies), showing that results \textit{predictably} do not differ.
    Finally, two experiments (forced choice and self-paced reading) are reported.
    The usage-based findings show a clear correlation with the behaviour of cognitive agents in both experiments.
    The present study therefore contributes to the well-established field of research into alternations using corpus and experimental methods.}
  \keywords{corpus methods and experimental methods, alternations, hierarchical models, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{Submitted April 2017. Do not quote.}
  \journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

For such probabilistic generalisations estimated on corpus data, there has always been an interest in correlating the findings with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.%
\footnote{In its colloquial meaning, the verb \textit{deduce} seems appropriate here.
However, in terms of the scientific method, we are most likely talking about \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
Technical problems of statistical inference aside, a successful experiment corroborates the theory, and a failed experiment weakens the theory.
The simplest assumption would be that the same approach is adequate for corpus studies.

@@

In this section, I introduce and illustrate the relevant alternating constructions.
I describe the narrowly defined syntactic configuration in which the alternation occurs and I motivate the focus on \textit{only} this narrow range rather than, for example, the whole range of nominal constructions expressing quantities.

@@

In the case at hand, three different syntactic configurations need to be distinguished with regard to case assignment inside German MNPs.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \citep{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

If the kind noun is bare -- \ie\ if it comes neither with a determiner nor a modifying adjective -- it has to agree in case with the measure noun as in (\ref{ex:intro:narrowapposition1}), and the genitive as seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition2}).

\begin{exe}
  \ex\label{ex:intro:narrowapposition}
  \begin{xlist}
    \ex[ ]{\label{ex:intro:narrowapposition1}\gll Wir trinken [[ein Glas]\Sub{Acc} [Wein]\Sub{Acc}]\Sub{Acc}.\\
    we drink a glass wine\\
    \glt We drink a glass of wine.}
    \ex[*]{\label{ex:intro:narrowapposition2} Wir trinken [[ein Glas]\Sub{Acc} [Weins]\Sub{Gen}]\Sub{Acc}.}
  \end{xlist}
\end{exe}

This construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.%
\footnote{The construction as in (\ref{ex:intro:narrowapposition1}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages in which the PGC with the synthetic genitive is not available.
This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to \textit{of} -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
Notice that the unavailability of the genitive on the kind noun can be seen as following from a rather quirky constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable (\citealp{GallmannLindauer1994,Schachtl1989}; see also \citealp[160]{Eisenberg2013b}).
%While only an accusative is shown in (\ref{ex:intro:narrowapposition1}), the kind noun obligatory agrees in case also with nominative and dative measure nouns.

The alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the PGC in (\ref{ex:intro:alternation1}) and the NAC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are by-and-large functionally and semantically equivalent.
However, Section~\ref{sec:analyses} is devoted to developing hypotheses about subtle differences between them.%
\footnote{Some descriptive and normative grammars take stronger positions with regard to the acceptability of the two options.
See \cite{Hentschel1993} and \cite{Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below (especially Section~\ref{sec:exp:fc}) should render it clear, there might be preferences, but we cannot assume either construction to be unacceptable.
}

@@

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP is summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb\ and the partitive genitive with a determiner in the kind NP the \PGCd.
For the alternants with an adjective but no determiner in the kind noun phrase I use the terms \NACa\ and \PGCa.
In principle, this paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.
However, in Section~\ref{sec:analyses}, the outer columns (\NACb\ and \PGCd) will still play a major role when the factors controlling the alternation are discussed.

I now turn to some more subtle issues related to the measure noun case alternation, namely:

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the \NACa\ and the \PGCa,}
%  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives,}
%  \item{\label{it:intro:femsg} case syncretism in feminine NPs,}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns,}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns, and}
  \item{\label{it:intro:other} alternative constructions for expressing quantities.}
\end{enumerate}

First of all, (\ref{it:intro:idiot}) refers to claims found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive (\PGCa) or case agreement (\NACa).
Overviews can be found in \cite{Hentschel1993} and \cite{Zimmer2015}.
Sentence (\ref{ex:intro:gennom}) shows a putative generic nominative on the kind noun inside an accusative MNP.

@@

It was shown empirically in \cite{Hentschel1993} that such sentences are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it virtually impossible to study them using corpus linguistic methods (Section~\ref{sec:corpusstudies}) and I consequently ignore them.

@@

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (794 of 861 cases, or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular.
%\footnote{More concretely, if these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear in their non-inflected form.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which have no normal plural forms at all.%
\footnote{Plurals like \textit{Pfunde} and \textit{Prozente} have special meanings and have very restricted uses, mostly in idiomatic expressions and prefabs such as \textit{Pfunde verlieren} `lose pounds' and \textit{Prozente machen} `make a profit'.
They cannot be used in normal MNPs.}
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected version \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected version \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected version denotes the quantity of wine fitting into two glasses, the inflected version denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, however, degrees of grammaticalisation as a factor influencing the alternation will be discussed prominently.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measuring quantities of substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morpho-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

Second, constructions with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' and \textit{mit} `with' are available, see (\ref{ex:alternatives}).

@@

These construction have very idiosyncratic properties.
The construction with \textit{voller} is discussed in \cite{Zeldes2018} and the construction with \textit{mit} is discussed in \cite{Bhatt1990}.
They are not semantically equivalent to the NAC and the PGC and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers, where the whole NP always refers to the container and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts and they are thus clearly not just more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined alternants.

@@

This section briefly reviews existing analyses of the \PGCa\ vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.
I discuss four main aspects:
first, the different syntactic structures of the alternants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (a preference to occur with cardinals);
fourth, preferences in different styles.

% MORPHO-SYNTAX AND PROTOTYPES

First, my analysis is based on the idea that in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, %the first question is which noun constitutes the head of the whole construction.
%This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
%Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
%Furthermore, the internal structure of MNPs can only be understood if the strong\slash weak inflection patterns of adjectives are taken into account.
the strong\slash weak inflection patterns of adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.%
\footnote{In the masculine and neuter singular genitive, the strong and weak forms are indistinguishable.
Since the alternation itself does not occur in the genitive (see Section~\ref{sec:descriptive}), this does not create any complications for the present study.}

\begin{exe}
  \ex\label{ex:starkschwach} 
  \begin{xlist}
    \ex{\label{ex:starkschwach:a}\gll mit ein\textbf{-em} stark\textbf{-en} Kaffee\\
                                      with a strong coffee\\}
    \ex{\label{ex:starkschwach:b}\gll mit stark\textbf{-em} Kaffee\\
                                      with strong coffee\\}
  \end{xlist}
\end{exe}

Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
On the one hand, they are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa.%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the PGC as a DP and in the NAC as an NP.}

>>>>

\documentclass[UKenglish]{article}

@@

\usepackage[small]{dgruyter}
\usepackage{microtype}
\usepackage{enumerate}
\usepackage{eqnarray}
\usepackage{colortbl}
\usepackage{soul}

@@

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer. Submitted April 2017. Do not quote.}
  %\affil[1]{}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{In this paper, a case alternation in German measure noun phrases is examined from the perspective of cognitively oriented corpus linguistics.
    In certain pseudo-partitive constructions, the embedded kind-denoting noun either agrees in its case with the measure head noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    My analysis is formulated in terms of Prototype Theory.
    I assume that the choice of alternants is influenced by non-alternating neighbouring constructions representing the prototypes more directly.
    I argue that the frequencies with which individual lemmas occur in these prototypical non-alternating constructions partially predict which alternant is chosen and that the genitive alternant is prototypical of strongly grammaticalised measure nouns.
    I present a large-scale corpus study using the DECOW corpus to support this theory.
    In the statistical analysis, I compare Maximum Likelihood estimators to Bayesian estimators (recently proposed for similar studies), showing that results \textit{predictably} do not differ.
    Finally, two experiments (forced choice and self-paced reading) are reported.
    The usage-based findings show a clear correlation with the behaviour of cognitive agents in both experiments.
    The present study therefore contributes to the well-established field of research into alternations using corpus and experimental methods.}
  \keywords{corpus methods and experimental methods, alternations, hierarchical models, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{Submitted April 2017. Do not quote.}
  \journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

For such probabilistic generalisations estimated on corpus data, there has always been an interest in correlating the findings with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.%
\footnote{In its colloquial meaning, the verb \textit{deduce} seems appropriate here.
However, in terms of the scientific method, we are most likely talking about \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
Technical problems of statistical inference aside, a successful experiment corroborates the theory, and a failed experiment weakens the theory.
The simplest assumption would be that the same approach is adequate for corpus studies.

@@

In this section, I introduce and illustrate the relevant alternating constructions.
I describe the narrowly defined syntactic configuration in which the alternation occurs, and I motivate the focus on \textit{only} this narrow range rather than, for example, the whole range of nominal constructions expressing quantities.

@@

\subsubsection{Core structures related to the alternation}

Three different syntactic configurations within MNPs need to be distinguished, and the case alternation occurs only in one of them.
It occurs only when the kind noun is modified by an attributive adjective and there is no determiner, as in (\ref{ex:intro:alternation}).
Superficially, the sentences are functionally and semantically equivalent either with the kind noun in the genitive (\ref{ex:intro:alternation1}) or in the same case as the measure noun, an accusative in the case of (\ref{ex:intro:alternation2}).%
\footnote{Some descriptive and normative grammars take stronger positions with regard to the acceptability of the two options.
See \cite{Hentschel1993} and \cite{Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As will be shown (especially in Section~\ref{sec:exp:fc}), there might be preferences, but we cannot assume either construction to be unacceptable.
}

This specific configuration has to be seen in the context of two other configurations, to which I turn now.
First, if the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \citep{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

Second, if the kind noun is bare -- \ie\ if it comes neither with a determiner nor a modifying adjective -- it is uninflected as in (\ref{ex:intro:narrowapposition1}), and the genitive as seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition2}).

\begin{exe}
  \ex\label{ex:intro:narrowapposition}
  \begin{xlist}
    \ex[ ]{\label{ex:intro:narrowapposition1}\gll Wir trinken [[ein Glas]\Sub{Acc} [Wein]\Sub{?}]\Sub{Acc}.\\
    we drink a glass wine\\
    \glt We drink a glass of wine.}
    \ex[*]{\label{ex:intro:narrowapposition2} Wir trinken [[ein Glas]\Sub{Acc} [Weins]\Sub{Gen}]\Sub{Acc}.}
  \end{xlist}
\end{exe}

This construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.%
\footnote{The construction as in (\ref{ex:intro:narrowapposition1}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages in which the PGC with the synthetic genitive is not available.
This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to \textit{of} -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
Notice that the unavailability of the genitive on the kind noun follows independently from a constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable (\citealp{GallmannLindauer1994,Schachtl1989}; see also \citealp[160]{Eisenberg2013b}).

It is difficult to determine whether the bare kind noun in the narrow apposition construction as in (\ref{ex:intro:narrowapposition1}) bears no case at all, a generic case, or agrees in case with the measure noun.%
\footnote{I would like to thank one of the reviewers for pointing this out.}
When there is an adjective as in (\ref{ex:intro:alternation2}), the embedded kind NP clearly agrees in case, but due to the overall absence of markers of case in the singular, bare nouns mostly show no indication of their case.
The only nouns which do have case markers in the singular are the so-called weak nouns \citep{Koepcke1995,Schaefer2016c}, which have something like a non-nominative \textit{-en} marker in the singular.
Unfortunately, there are no genuine mass nouns among the weak nouns.
However, a few of them can be coerced into a mass noun, such as \textit{Hase} `rabbit' (meaning `rabbit meat').
It then appears as if the uninflected form is preferred, but the inflected form is not excluded.
In (\ref{ex:macabre1}), the clearly acceptable form \textit{Hase} can only be a nominative singular or caseless.
In (\ref{ex:macabre2}), the form \textit{Hasen} could be an accusative, dative, or genitive.

\begin{exe}
  \ex\label{ex:macabre}
  \begin{xlist}
    \ex[ ]{\label{ex:macabre1}\gll Niemand will [ein Stück [Hase]\Sub{Nom\slash caseless}]\Sub{Acc} essen.\\
    nobody wants a piece rabbit eat\\
  \trans Nobody wants to eat a piece of rabbit.}
    \ex[?]{\label{ex:macabre2}\gll Niemand will [ein Stück [Hasen]\Sub{Acc\slash Dat\slash Gen}]\Sub{Acc} essen.\\
    nobody wants a piece rabbit eat\\}
  \end{xlist}
\end{exe}

Even a full unacceptability of (\ref{ex:macabre2}) would not be conclusive, however, as a possible aversion of speakers towards the case-marked form might be due to the fact that it is at least potentially also a genitive, in which case the constraint against bare genitive nouns would apply.
With plural kind nouns, the obligatory marking of the dative plural with \textit{-en} might provide some clues.
However, plural kind nouns do not behave like singular ones in measure phrases anyway, as will be argued in Section~\ref{sec:neighbouringcases}.
Also, judgements vary between (\ref{ex:datpl1}) and (\ref{ex:datpl2}), and both are found in corpora.%
\footnote{For example, the variant in (\ref{ex:datpl1}) with the lemmas \textit{Sack} and \textit{Apfel} occurs four times, and the one in (\ref{ex:datpl2}) twice in the very large DECOW corpus (see Section~\ref{sec:gettingdata}).
Similarly, for [\textit{Kiste} [\textit{Äpfel}]\Sub{Nom\slash Acc\slash Gen}]\Sub{Dat} `box of apples' (no case identity), I find three examples, and [\textit{Kiste} [\textit{Äpfeln}]\Sub{Dat}]\Sub{Dat} (clearly marked case identity), I find six examples.
}

\begin{exe}
  \ex\label{ex:datpl} 
  \begin{xlist}
  \ex{\label{ex:datpl1}\gll mit [zwei Säcken [Äpfel]\Sub{Nom\slash Acc\slash Gen}]\Sub{Dat}\\
     with a sack apples\\
    \trans with a sack of apples}
    \ex{\label{ex:datpl2}\gll mit [zwei Säcken [Äpfeln]\Sub{Dat}]\Sub{Dat}\\
    with a sack apples\\}
  \end{xlist}
\end{exe}

Descriptive grammars seem to favour an analysis in terms of caselessness (for example, \citealp[1981]{ZifonunEa1997c}).
The hazy picture of the case of bare kind NPs is most likely due to the fact that case is so rarely marked on German nouns, where case is marked mostly on determiners and to some degree adjectives.
The uncertainty in the few cases where case can be marked (weak nouns in the singular and dative plurals) would thus be a direct consequence of the fact that the construction is not very specific with respect to the case of the kind noun.

@@

To summarise, the case patterns in the NAC and in the PGC (depending on the structure of the kind NP) is given in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb\ and the partitive genitive with a determiner in the kind NP the \PGCd.
For the alternants with an adjective but no determiner in the kind noun phrase I use the terms \NACa\ and \PGCa.
In principle, this paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.
However, in Section~\ref{sec:analyses}, the outer columns (\NACb\ and \PGCd) will still play a major role when the factors controlling the alternation are discussed.

\subsubsection{Neighbouring cases}
\label{sec:neighbouringcases}

% namely:
% 
% \begin{enumerate}[i.]
%   \item{\label{it:intro:idiot} alleged alternatives to the \NACa\ and the \PGCa,}
% %  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives,}
% %  \item{\label{it:intro:femsg} case syncretism in feminine NPs,}
%   \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns,}
%   \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns, and}
%   \item{\label{it:intro:other} alternative constructions for expressing quantities.}
% \end{enumerate}
%
% \vspace{-1\baselineskip}

I now turn to some more subtle issues related to the measure noun case alternation in order to delimit the scope of the study.
First, there is a claim found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive (\PGCa) or case agreement (\NACa).
Overviews can be found in \cite{Hentschel1993} and \cite{Zimmer2015}.
Sentence (\ref{ex:intro:gennom}) shows a putative generic nominative on the kind noun inside an accusative MNP.

@@

It was shown empirically in \cite{Hentschel1993} that such sentences are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it virtually impossible to study them using corpus linguistic methods (Section~\ref{sec:corpusstudies}), and I consequently do not discuss them further.

@@

Second, we find that, if the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (794 of 861 cases, or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular.
%\footnote{More concretely, if these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

Third, some measure nouns have been grammaticalised in a way that they always appear in their non-inflected form.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which have no normal plural forms at all.%
\footnote{Plurals like \textit{Pfunde} and \textit{Prozente} have special meanings and have very restricted uses, mostly in idiomatic expressions such as \textit{Pfunde verlieren} `lose pounds' and \textit{Prozente machen} `make a profit'.
They cannot be used in normal MNPs.}
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected version \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected version \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected version denotes the quantity of wine fitting into two glasses, the inflected version denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, however, degrees of grammaticalisation as a factor influencing the alternation will be discussed prominently.

@@

Finally, there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measurement of quantities of substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morpho-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
There is an analytic (pseudo-)partitive with \textit{von} `of'.
It is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

Additionally, we find constructions with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' and \textit{mit} `with' are available, see (\ref{ex:alternatives}).

@@

These construction have very idiosyncratic properties.
The construction with \textit{voller} is discussed in \cite{Zeldes2018} and the construction with \textit{mit} in \cite{Bhatt1990}.
They are not semantically equivalent to the NAC and the PGC, and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers, where the whole NP typically refers to the container and never to the quantity contained in it.
They are incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts, and they are thus clearly not just more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined alternants.

@@

This section briefly reviews existing analyses of the \PGCa\ vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

.%
\footnote{A reviewer mentioned that she or he had the impression that dialectal variation is also a factor influencing the alternation.
This is definitely true as some dialects (such as Alemannic) tend to have no genitive at all.
While this is an interesting aspect for further research, the main obstacle is that there are no corpora of German which are both large enough and annotated with reliable meta data about regional variants.
In the annotation of the corpus study, documents obviously written in a regional variant were excluded.
}
